\subtopic{Helium - Neon Laser}

\begin{itemize}
	\item It is a 4 level Gaseous system.
	\item Created by Ali Jawan, W. Bennet, D. Harrul in 1962
	\item It produces continuous output.
	\item Pumping mechanism: 10kV electron dischange
	\item Active medium: mixture of He and Ne in 10:1 ratio. (Active center is Ne)
	\item wavelength of light emitted: 632.8 nm (red color)
\end{itemize}

\input{../images/he-ne-hw.tex}

\ssubsection{Working}
Due to electric dischange in gas, energetic electron interacts with ground state helium atoms. The impact of $e^{-}$ results in exchange of some of its energy to the atom. As a result. Helium atoms gets excited to higher energy levels: $F_{1} \& F_{2}$. These two energy levals are close to $E_{4} \& E_{6}$ levels of Neon atoms and collision of second kind takes place between them, hence we goes to excited state. $E_{5}$ of $E_{3}$. $E_2$ level of we forms metastable state.

3 types of transition are possible :

\begin{itemize}
	\item $E_4 \rightarrow E_3$, emitting infrared light of $\lambda=1.15 \mu \mathrm{m}$
	\item $E_6$ $\rightarrow E_5$, emitting $F R$ of $\lambda = 3.39 \mu \mathrm{m}$
	\item $E_6 \rightarrow E_3$, emitting red light of $\lambda = 632.8$ nm
\end{itemize}

IR light emitted is absorped by the mirrors. Neon atoms in terminal level $E_{3}$ decay very rapialy to $E_{2}$ metastable state, much faster than spontaneously decaying from $E_{6} \rightarrow E_{3}$ level. Thus lower level $E_3$ is kept empty and population inversion is achieved between $E_{6} \& E_{3}$. Intensity of light produced is inversely proportional to diameter of quartz tube.
\bigskip

\resizebox{1\textwidth}{!}{
	\input{../images/he-ne.tex}
}